---
title: Cultural evolution and software development
author: Pierce Edmiston
output:
  beamer_presentation:
    theme: "metropolis"
    latex_engine: "lua"
---
```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(crotchet)

opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE)
sapply(list.files("src/R", "*.R", full.names = TRUE), read_chunk)
sapply(list.files("src/dot", "*.gv", full.names = TRUE), read_graphviz_chunk)
theme_set(theme_minimal(base_size = 20))
```
```{r peaks, include=FALSE}
```

# About me

Studied cultural evolution in the Psychology Dept at UW-Madison.  
Starting thinking about software development as a "model organism".  
Hoping to convince you that **legacy code isn't as bad as you think.**

<!--
Today I'm going to talk about cultural evolution and software development.
Cultural evolution is a topic I studied in grad school and as a postdoc, and as
I learned more about more about it, I became really interested in how the ideas
and debates about cultural evolution could be studied by looking at software
development. I started to think about software development as a sort of model
organism for studying cultural evolution. That perspective has really changed
how I view software development, and I hope to share some of that perspective
with you today. Ultimately, what I discovered is that **legacy code isn't as bad
as we think.** Now I'm certainly not going to say that all legacy code is good
and we should never write new software. But ultimately I think that we as
crafters are far better at handing off codebases from one generation to the next
than we give ourselves credit for.
-->

# How long does it take to run 400 meters?

```{r relay-races, out.height="80%", fig.cap="World record sprint times. Data from Wikipedia."}
```

<!--
To start I want to ask a question about relay races. How long does it take to
run 400 meters? The world record holder for the 400m is Wayde van Niekerk from
South Africa who in 2016 ran 400 meters in 43.03 seconds. This amounts to a bit
under 11 seconds per 100 meters. But Usain Bolt, the world record holder for
the 100m, can run 100 meters in 9.58 seconds. So a faster way to cover 400
meters might be to have Usain Bolt run 100 meters 4 separate times, which would
add up to a total of 38.32 seconds. This would be faster than van Niekerk, but
it turns out that the fastest way to run 400 meters is as a relay team. In 2012
the Jamaican team, which included Bolt, moved a baton 400 meters in 36.84
seconds, 6 seconds faster than van Niekerk and 1.5 seconds faster than four
times Bolt's fastest 100 meter.
-->

# Why is a relay strategy effective for running races?

Each runner has less distance to run.  
Only one runner has to accelerate from zero.  
**Each person gets a head start on running.**

<!--
Why is a relay strategy effective for running races? Part of the reason
is that each person has a shorter distance to run, and is less likely to
get fatigued. But in order to explain why a relay team can be faster than
the fastest individual, we need an additional explanation. In this case,
the reason is that sprinters are at their slowest when accelerating, and
usually take between 30 and 50 meters to reach peak velocity. In a relay
team, only one person has to accelerate from 0, and everyone else receives
the baton while already in motion. Part of the reason that a relay strategy
is effective for running races is that each person gets a head start
on running.
-->

# Is a relay strategy effective for problem solving?

The baton is a tool or other solution to a problem.  
Passing the baton would seem to give a head start, but how much?  
**How do we know whether a relay strategy is effective?**

\vspace{1em}

```{r team-structures, out.height="55%", fig.align="left"}
crotchet::draw_graphviz("team-structures", package = "totems")
```

<!--
The question I was after in grad school is whether a relay strategy can ever
similarly effective for problem solving. If we think about the baton as a tool
or some other solution to a problem, we can pass that tool on to someone else
to give them a head start, but how much does it really help them? How do we
know whether a relay strategy is effective?

My answer to this question is summarized in this diagram. The figure on top
is the relay strategy: one person starts on a solution to a problem before
handing it off to the next generation. I call this "diachronic collaboration"
because it's teamwork that extends over time. We can compare this strategy
to "synchronic collaboration" where the teamwork is happening at the same time.
Finally, we have individual work, where one person "goes it alone".

My argument was basically that if we ever compare these three strategies and
find that the relay strategy is the most effective, then we are forced to
conclude that a relay strategy does more than just give you a head start
on a problem, it actually changes the way you solve it. It makes you more
effective than you would have been if you would have tried to do it alone.
-->

# Is teamwork effective for problem solving?

Surprisingly, "assembly bonus effects" are rare in psychology.  

- Collective memory
- **Wason selection task**

<!--
Before we answer this question about relay teams, it's worth taking a step back
and asking whether traditional, synchronic teamwork helps with problem solving.
These effects are referred to as "assembly bonus effects", where the sum is
greater than its parts. Surprisingly, these assembly bonus effects are
actually pretty rare in the psychological literature. Working in a team on
a problem often doesn't raise the ability of the team beyond the most effective
member. There are a few cases that have been documented. One example is on
collective memory, where if I were to flash up a screen full of objects
and then ask you questions about them, you would be better off working in a
team as opposed to as an individual. Another example is the Wason selection
task. This is a little logic puzzle that's pretty easy to demo, so let's give
it a shot.
-->

# Wason selection task

Each card has a number on one side, and a patch of color on the other.  
> If a card is even on one side, then it's red on the other.  
**Which cards do we need to flip over to prove it?**

```{r wason-selection-task, fig.cap="Wason selection task, from Wikipedia."}
include_graphics("img/wason.pdf")
```

<!--
/poll "Which cards do we need to flip over to prove the rule?" "AB" "AC" "AD" "BC" "BD" "CD" anonymous
-->

# Wason selection task

Each card has an age on one side, and a beverage on the other.  
> If a person is under 21, they should be drinking soda.  
**Which two cards do we need to flip to prove it?**

```{r wason-selection-task--social, fig.cap="Wason selection task, from Wikipedia."}
include_graphics("img/wason-social.pdf")
```

<!--
-->

# Teamwork is good for exploration / exploitation

```{r wildcat-wells, fig.cap="Wildcat Wells experiment (Mason \\& Watts, 2012, ``Collaborative learning in networks'').", out.height="80%"}
gridExtra::grid.arrange(
  crotchet::read_image("img/wildcat-wells-fitness.png"),
  crotchet::read_image("img/wildcat-wells.png"),
  ncol = 2
)
```

<!--
-->

# When is a relay strategy effective for problem solving?

```{r diachronic-collaboration, engine="dot"}
```

1. A relay strategy can remediate the costs of fixation.
1. A relay team with complementary skills is "greater than its parts."

# Solving a problem as climbing a hill

A problem solver exists as a point in the landscape.  
They can see some distance in X and Y.  
Each time step they try to move up the hill.  

\vspace{1em}

```{r simple-hill, out.height="60%", fig.align="center"}
```

# Visualizing complementary skills

```{r ability-as-vision, out.height="90%", out.width="80%", fig.height=10, fig.width=8}
```

<!--
Agents have skill corresponding to vision in two dimensions.
Agents have equal or unequal skill along those two dimensions.
-->

# Results

```{r differing-skills, fig.height=10, fig.width=10, out.width="80%", out.height="80%"}
```

# Explanation

```{r differing-skills-walk}
gg_differing_skills_walk
```

# Complementary codebases

**Proposal:** Rebrand legacy codebases as ``complementary.''  
Is this why solving problems in different languages is so valuable?  

# Real teams blend all forms of teamwork

```{r tina-team, engine="dot"}
```

# Human culture as a really long relay race

Why are humans effective problem solvers?

- We evolved specialized intelligences.
- **We developed systems of cultural inheritance**.

<!--
The reason this question is important is because of it's relevance to ongoing
debates about human evolution, in particular, what it is that makes us such
effective problem solvers as compared to other animals. If I had to simplify the
debate somewhat, I would say that on one side we have people saying that we
evolved specialized intelligences that enable advanced reasoning processes that
improve individual problem solving abilities.

An alternative view is that we owe our success in problem solving in large
part to our systems of cultural inheritance. On this view, it's not our
ability to solve problems as individuals that is all that impressive. Instead
it's our ability to copy other people that makes us adaptive problem solvers.
Humans may be smart, but cultures are smarter.
-->

# Lost European explorer experiments

> [S]tarvation on nardoo is by no means unpleasant, but for the weakness one
  feels, and the utter inability to move oneself, for as the appetite is
  concerned, it gives me the greatest satisfaction.

```{r nardoo, out.height="40%", fig.cap="The nardoo aquatic fern, native to Australia."}
crotchet::draw_image("img/nardoo.jpg")
```
  
<!--
A favorite source of evidence for why cultures are more adaptive than
individuals comes from what has been termed the Lost European explorer
experiments. This is where a group of intelligent and well-equipped adventurers
somehow get stranded in a foreign land, and the test is whether they can
survive off the land long enough to make it back home. The irony of these
experiments is that the explorers often die right alongside indigenous cultures
who have somehow figured out how to thrive in the same environment.

An example of this is the Burke and Wills expedition across Australia. As
they ventured across inland Australia they were running out of supplies, but
they ran into a group of native Australians that were surviving off the land just
fine. They saw that these Australians survived largely by eating an aquatic
fern called nardoo. The Australians prepared this plant by mashing it into
a paste and forming it into little cakes. But all the explorers saw was that the
Aborigines were eating a lot of nardoo, so they assumed it wasn't harmful. It
turns out that the food preparation process removed toxins from the nardoo,
and when eaten unprepared, it caused a form of thamine deficiency that ultimately
lead to their deaths.

Here is a passage from Wills diary shortly before he died, describing how
odd it was to die of starvation with a full belly.

> [S]tarvation on nardoo is by no means unpleasant, but for the weakness one
  feels, and the utter inability to move oneself, for as the appetite is
  concerned, it gives me the greatest satisfaction.

The Lost European explorer experiments demonstrate that cultures seem to be
better at solving the problems necessary for survival than any individual
plopped down in the same environmental conditions.
-->

# Lost Software Crafter experiments

- Trying to follow an out-of-date README to set up a dev machine.
- Googling a problem that may or may not be related to my IDE.
- Completing an 8th Light apprenticeship solely from the syllabus.

\vspace{2em}

> Running these long bash commands is by no means unpleasant, but for the errors
  they generate, and the utter inability to serve the app at localhost:8080,
  for as my mechanical keyboard is concerned, it gives me the greatest
  satisfaction.

<!--
Looking back at the Burk and Wills expedition it made me think of what the
analogous situation might be at 8th Light. I can imagine some "Lost Software
Crafter experiments" that may sound familiar to some of you, like
"Trying to follow an out-of-date README to set up a dev machine."
"Googling a problem that may or may not be related to my IDE."
"Completing an 8th Light apprenticeship solely from the syllabus."
I can imagine writing something like this in my journal:

> Running these long bash commands is by no means unpleasant, but for the errors
  they generate, and the utter inability to serve the app at localhost:8080,
  for as my mechanical keyboard is concerned, it gives me the greatest
  satisfaction.
-->

# Crafters are experts at cultural inheritance.

Programming languages were invented for humans.  
Open source software removes a huge barrier to entry.  
DevOps practices make it easier for the next generation.  

<!--
-->

# Why do crafters care about tooling?

```bash
# dp-toolchain
dp ssm tina/dev
```

<!--

-->

# Why legacy code isn't as bad as you think

Working software is a valuable cultural artifact.  
Rebrand it as ``complementary code''!  

# Cultural evolution and software development

Pierce Edmiston  
[github.com/pedmiston/miniconf21](https://github.com/pedmiston/miniconf21)
